Where Do Events Come From?

### 2018-09-27 paper.tex
(>) suggestions from reviews

\subsection{Scoped Interrupt Service Routines}
\label{sec.isrs}

@@

{ pinMode(13, OUTPUT); }
output (high/low v) PIN_13 do
    { digitalWrite(13, @v); }
end

@@

input high/low PIN_02;
{
    pinMode(2, INPUT_PULLUP);
    EICRA |= (1 << ISC00);  // sets INT0
    EIMSK |= (1 << INT0);   // turns on INT0
}
spawn async/isr [INT0_vect] do
    emit PIN_02({digitalRead(2)});
end
\end{lstlisting}

Listing~\ref{lst.gpio.async}, the asynchronous side, implements the driver for
the output and input events.
%
An \code{output} implementation (ln~4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.
\CEU supports inline C between curly braces (ln~3,5) with interpolation to
evaluate \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries of embedded
toolchains such as Arduino.
%
In the example, when the driver is included, it configures pin 13 as output
(ln~3) and sets its new state whenever \code{PIN\_13} is emitted (ln~5).
%
An \code{input} event implementation uses an ISR registered with the
\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will typically perform simple operations and emit an input event
to awake the synchronous side (ln~17).
However, although the ISR executes asynchronously when the interrupt occurs,
the \code{emit} goes into a queue and does not affect the synchronous side
immediately.
%
In the example, when the input driver is included, it also configures pin 2 to
behave as input and to generate external interrupts on level transitions
(ln~11--15).

@@

The addition of asynchronous ISRs now poses real threats concerning race
conditions since they interrupt programs at arbitrary points and could possibly
share variables with synchronous code.
%
Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
%
The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
and a buffer to prevent data loss (ln~4).
The buffer is indispensable because the microcontroller might not cope with the USART 
speed.
%
The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
buffer (ln~9) and signals the application (ln~10).
%
As a possible race condition, note that the ISR may fire and append a new byte
to the buffer (ln~9) just before the application clears it (ln~21), in which
case the new byte will be lost forever.

\CEU treats a \code{spawn~async/isr} as a block of code that runs in parallel with
the rest of the program (hence the prefix \code{spawn}).
%
This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
Listing~\ref{lst.usart} may lead to race conditions (e.g., ln~9,21), and \CEU
raises a compile-time error.
%
The programmer is responsible to protect concurrent memory accesses with
\code{atomic} blocks, which disables interrupts for a short period
of time.

@@

spawn async/isr [USART_RX_vect] do
    rx_buf = rx_buf .. [{UDR0}];  // appends new byte
    if $rx_buf == 1 then
        emit USART_RX; // emits unless already pending
    end                // (assumes client consumes all)
end

@@

However, we do not expect that application programmers should be required to
resolve race conditions.
\CEU supports reactive abstractions (similar to coroutines) that help hiding
concurrency issues inside drivers.
%
Listing~\ref{lst.usart.code} changes the USART API (ln 1--3) to expose a code
abstraction that expects a reference to a buffer and a number of bytes to
receive.
Now, the application (ln~29--35) invokes the abstraction (ln~33) passing a
local buffer (ln~32).
%
The driver (ln~5--27) now hides the low level interface of the previous example
(ln~7--8) and implements the code abstraction (ln~15--27) that protects the
accesses to the shared buffer with an \code{atomic} block (ln~18--21).
%
The abstraction copies the driver buffer into the application buffer (ln~19)
up to the requested number of bytes (ln~22).
%
On the one hand the extra copying incurs a memory and runtime overhead, but on
the other hand it prevents race conditions with the ISR.

>>>>

\section{Interrupt Service Routines in \CEU}
\label{sec.contributions}

In Listing~\ref{lst.ceu}, the input \& output events are external to the
program and are generated by the host language representing the environment.
%
We now propose to extend \CEU with interrupt services routines (ISRs) to take
control of the whole event loop using a safe and energy-efficient approach.

\subsection{Scoped ISRs}
\label{sec.isrs}

@@

{ pinMode(13, OUTPUT); }
output (high/low v) PIN_13 do
    { digitalWrite(13, @(v as int)); }
end

@@

input high/low PIN_02;
{
    pinMode(2, INPUT_PULLUP);
    EIMSK |= (1 << INT0);  // enables pin-02 ints
    EICRA |= (1 << ISC00); // for level changes
}
spawn async/isr [INT0_vect] do
    emit PIN_02({digitalRead(2)} as high/low);
end
\end{lstlisting}

Listing~\ref{lst.gpio.async}, the asynchronous side, implements the driver for
the output and input events.
%
An \code{output} implementation (ln~4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.

\CEU is designed to interoperate seamlessly with C and supports a compatible
type system with self-evident conversion rules and same internal representation
(e.g., \code{high/low} is a boolean, \code{u8} is an unsigned 8-bit value,
etc.).
\CEU also supports inline C between curly braces (ln~3,5) and interpolation to
evaluate \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries of embedded
toolchains such as Arduino.
Our drivers still use C, but only a convenient subset (e.g., assignments and
getters \& setters) for low-level port manipulation.

In the example, when the driver is included, it configures pin 13 as output
(ln~3) and sets its new state whenever \code{PIN\_13} is emitted (ln~5).
%
An \code{input} event implementation uses an ISR registered with the
\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will typically perform simple operations and emit an input event
to awake the synchronous side (ln~17).
However, although the ISR executes asynchronously when the interrupt occurs,
the \code{emit} goes into an input queue and does not affect the synchronous
side immediately.
%
The input driver also configures pin 2 to behave as input and to generate
external interrupts on level transitions (ln~11--15).
%
The queue is static and new emissions of an unhandled input override the same
position in the queue.
This semantics is comparable to interrupt flags in microcontrollers and
prevents buffer overflows.

@@

The addition of asynchronous ISRs now poses real threats concerning race
conditions since they interrupt programs at arbitrary points and could possibly
share variables with synchronous code.
%
Listing~\ref{lst.usart} illustrates a minimum serial communication (USART)
application (ln~15--24) that consumes incoming bytes (ln~19--23) as they arrive
(ln~18).
%
The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
and a buffer to prevent data loss (ln~4).
The buffer is indispensable because the microcontroller might not cope with the USART 
speed.
%
The (unsafe) driver ISR (ln~6--13) simply appends incoming bytes to the end of the
buffer (ln~9) and signals the application (ln~11).
%
As a possible race condition, note that the ISR may fire and append a new byte
to the buffer (ln~9) just before the application clears it (ln~23), in which
case the new byte will be lost forever.

\CEU treats a \code{spawn~async/isr} as a block of code that runs in parallel with
the rest of the program (hence the prefix \code{spawn}).
%
This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
Listing~\ref{lst.usart} may lead to race conditions (e.g., ln~9,23), and \CEU
raises a compile-time error.
%
The programmer is responsible to protect concurrent memory accesses with
\code{atomic} blocks, which disables interrupts for a short period
of time.

@@

spawn async/isr [USART_RX_vect] do
    rx_buf = rx_buf .. [{UDR0} as byte];  // appends
    if $rx_buf == 1 then
        emit USART_RX; // emits unless already pending
    end                // (assumes client consumes all)
end

@@

However, we do not expect that application programmers should be required to
resolve race conditions.
\CEU supports \code{code} reactive abstractions (similar to coroutines) that help hiding
concurrency issues inside drivers.
%
Listing~\ref{lst.usart.code} changes the USART API (ln 1--3) to expose a code
abstraction that expects a reference to a buffer and a number of bytes to
receive.
Now, the application (ln~29--35) invokes the abstraction (ln~33) passing a
local buffer (ln~32).
%
The driver (ln~5--27) now hides the low level interface of the previous example
(ln~7--8) and implements the code abstraction (ln~15--27) that protects the
accesses to the shared buffer with an \code{atomic} block (ln~18--21).
%
The abstraction copies the driver buffer into the application buffer (ln~19)
up to the requested number of bytes (ln~22).
%
On the one hand the extra copying incurs a memory and runtime overhead, but on
the other hand it prevents race conditions with the ISR.
